A Large-Scale Study of the Use of Eval in R

### 2020-08-28 inst/analysis/analysis_stable.Rmd
Consistence

---
title: "Empirical Evaluation of eval in R"
output:
  rmdformats::readthedown:
        code_folding: hide
        lightbox: true
        gallery: false
        df_print: paged
        toc_depth: 3
date: "`r Sys.Date()`"
editor_options: 
  chunk_output_type: console
params:
  base_dir: /var/lib/R/project-evalR

@@

```{r subsample_for_ast}
#ast_sizes <- eval_calls %>% slice_sample(n=10000) %>% mutate(expr_resolved_nodes = map_int(expr_resolved, expr_size_str))
# Now precomputed during the tracing
ast_sizes <- read_fst(path(RUN_DIR, "package-evals-traced.3/summarized-ast-size.fst")) %>% as_tibble() %>% 
      filter(eval_source_type != "<undefined>")
```

@@

ast_sizes %>%
  ggplot(aes(x = eval_source_type, y = expr_resolved_nodes, fill = eval_source_type, weight = nb_ev_calls), na.rm = TRUE) +
  geom_violin(trim = FALSE, width=.5)+
  geom_boxplot(width=.05) +
  scale_y_log10(labels = scales::comma) +
  labs(y="AST size (log)") +
  theme(
    legend.position=c(0.5, 0.85),
    legend.box.background = element_rect(fill="white", size=0.1),
    legend.box="horizontal",
    axis.title.x=element_blank()
  )

@@

```{r}
eval_calls %>% semi_join(expr_resolved_polymorphism %>% filter(polym_degree == max(polym_degree)), by ="eval_call_srcref")
```

>>>>

---
title: "Empirical Evaluation of eval in R"
output:
  rmdformats::readthedown:
        code_folding: hide
        lightbox: true
        gallery: false
        df_print: paged
        toc_depth: 3
date: "`r Sys.Date()`"
editor_options: 
  chunk_output_type: console
params:
  #base_dir: /home/pierre/Documents/Articles/ISCE2021
  base_dir: /var/lib/R/project-evalR

@@

```{r subsample_for_ast}
#ast_sizes <- eval_calls %>% slice_sample(n=10000) %>% mutate(expr_resolved_nodes = map_int(expr_resolved, expr_size_str))
# Now precomputed during the tracing
ast_sizes <- read_fst(path(RUN_DIR, "package-evals-traced.3/summarized-ast-size.fst")) %>% as_tibble() %>% 
      filter(eval_source_type != "<undefined>") %>% 
      rename(dataset = eval_source_type)
```

@@

ast_sizes %>%
  ggplot(aes(x = dataset, y = expr_resolved_nodes, fill = dataset, weight = nb_ev_calls), na.rm = TRUE) +
  geom_violin(trim = FALSE, width=.5)+
  geom_boxplot(data = slice_sample(ast_sizes, n = 100000), width=.03) + # too slow otherwise
  scale_y_log10(labels = scales::comma) +
  labs(y="AST size (log)") +
  theme(
    legend.position=c(0.5, 0.85),
    legend.box.background = element_rect(fill="white", size=0.1),
    legend.box="horizontal",
    axis.title.x=element_blank()
  )

@@

The maximum number of types is `r max(expr_resolved_polymorphism$polym_degree)`.

```{r}
highly_polymorphic <- eval_calls %>% semi_join(expr_resolved_polymorphism %>% filter(polym_degree == max(polym_degree)), by ="eval_call_srcref")
```

```{r}
highly_polymorphic %>% count(caller_function, wt = nb_ev_calls, sort = TRUE)
```

### 2020-08-28 inst/paper/main.tex
Consistence

Richards et al.~\cite{ecoop11} provided the first large-scale study of the
runtime behavior of \eval in JavaScript. They analyzed a corpus of the
10,000 most popular websites. They used an instrumented web browser to
gather execution traces and perform a dynamic analysis.  Their results show
that \eval is pervasive with 82\% of the most popular websites using it. The
reasons for using \eval include the desire to load code on demand (to reduce
the time it takes to display a page), the deserialization of JSON data
(about 37\% of all calls) and lightweight meta-programming to customize web
pages.  While many uses \eval were legitimate, many were unnecessary and
could be replaced with equivalent and safer code.  The reported on a
categorization of inputs to \eval that covers the vast majority of input
strings.  Restricting themselves to \eval in which all named variables refer
to the global scope, many patterns can be replaced by more disciplined
code~\cite{oopsla12b}.  The limitation of that work are that there was no
measure of code coverage, so the numbers presented can be a lower bound on
the possible behaviors.  Furthermore in 2011, JavaScript usage is quite
different from today. Their work captured very few libraries and none of the
more recent Node.js ecosystem.  More details about dynamic analysis of
JavaScript can be found in~\cite{liang}.

Wang et al.~\cite{wang} analyzed usage of dynamic features in 18 Python
programs include dynamic code generation to find if they affects file
change-proneness.  They reported that files with dynamic features are
significantly more likely to be the subject of changes, than other files.
Chen et al. looked at the correlation between code changes and dynamic
features, including \eval, in 17 Python programs~\cite{chen}. They did not
observe many uses of \eval.  Callau et al. performed an empirical study of
the usage of dynamic features in the 1,000 largest Smalltalk projects in the
Squeaksource repository, accounting for more than 4 million lines of
code~\cite{oscar}. While \eval itself is not present, Smalltalk has a rich
reflective interface that gives programmer great flexibility. Their work was
based on a static analysis.

Morandat et al.~\cite{ecoop12} had a short section on the usage of \eval in
R. They found the \eval function to be widely used in R code with 8500 call
sites in CRAN and 2.2 million dynamic calls. The 15 most frequent call sites
account for 88\% of eval. The \c{match.arg} function is the highest user
with 54\% of all calls. In the other call sites, they saw two uses
cases. The most common is the evaluation of the source code of a promise
retrieved by \c{substitute} in a new environment. This is done in the
\c{with} function. The other use case is the invocation of a function whose
name or arguments are determined dynamically. For this purpose, R provides
\c{do.call} and thus using \eval is overkill.

@@

The R Project is a tool for data analysis.  At its heart is a {vectorized,
  dynamic, lazy, functional, object-oriented} programming language with a
rather unusual combination of features~\cite{ecoop12} designed to ease
learning by non-programmer and enable rapid development of statistical
methods~\cite{R96}.  The language was designed as a successor to
S~\cite{S88}.

In R, most data types are vectorized. Values are constructed by the
\c{c(...)} function: \c{c("hi","ho")} creates a vector of two strings. The
language does not differentiate scalars from vectors. To enable equational
reasoning, R copies values accessible through multiple variables when they
are written to.  Values can be tagged by user-defined attributes. For
instance, one can attach the attribute \c{dim} to the value
\c{x<-c(1,2,3,4)} by evaluating \c{attr(x,"dim")<-c(2,2)}.  Once done,
arithmetic functions will treat \c x as a \c{2x2} matrix. Another attribute
is \c{class} which can be bound to a list of names. For instance,
\c{class(x)<-"human"}, sets the class of \c{x} to \c{human}.  Attributes are
used for object-oriented dispatch.  Every linguistic construct is desugared
to a function call, even control flow statements, assignments, and
bracketing. Furthermore, all functions can be redefined in user code. This
makes R both flexible and challenging to compile~\cite{dls19}.

Arguments to user-defined functions are bundled into a thunk called a
promise. Logically, a promise combines an expression's code, its
environment, and its value.  To access the value of a promise, one must
force it. Forcing a promise triggers evaluation and the computed value is
captured for future reference.

R supports reflection and meta-programming. The \c{substitute(e,envir)}
function yields the parse tree of the expression \c{e} after performing
substitutions defined by the bindings in \c{envir}.

\begin{lstlisting}
> substitute(expression(a + b), list(a = 1)))  
expression(1 + b)
\end{lstlisting}
\vspace{2mm}

\noindent
R allows programmatic manipulation of parse trees, which are themselves
first class objects. They are evaluated using the \c{eval(e,envir)}
function. Environment on the call stack can also be accessed.

@@

\begin{lstlisting}
> r <- 5
> env <- new.env()
> env\$r <- 9
> eval(r, envir = env)
[1] 5
> evalq(r, envir = env)
[1] 9
\end{lstlisting}

@@

To roughly estimate the complexity of the code being evaluated,
Fig.~\ref{fig:ast-size} shows the distribution of the abstract syntax tree
size of the \c{expr} argument to \eval. The majority of inputs across core
and packages (\NbAstOnePercent of \eval calls and \NbAstOneCallSitePercent of call sites) is 1. That is to say, the code being evaluated
consists of a single variable name or constant. A function call needs at
least three nodes. Inputs smaller than \AstSizeNineFive nodes account for 95\% of all
calls. Sizes are roughly similar between core and packages.

@@

One of the main differences of the R \eval with the javascript one is how it
interacts with \emph{scope}, or environments in R. In Javascript, \c{eval} can
access the local scope and the global scope. In R, \c{eval} has an argument
\c{envir} to indicate in which environment its first argument must be evaluated.
By default, it is equal to \c{parent.frame()}, 
which is the parent environment of \c{eval} call.This default argument
happens in \DefaultEnvirExprPercent of the eval calls.

@@

We looked at the environments passed to all the \eval calls in our corpus.
Table~\ref{tab:environments} summarizes the results. A numeric environment class
\c{n} denotes the environment of the $n$-th call stack frame from the current
function. \c{global} denotes the top-level environment and \c{list} denotes a list
passed for evaluation of formulas. Environment classes of the form $n+$ denote
the $n$-th environment extended with a new environment. The new environment
provides a limited form of sandboxing. All assignments using the \c{\<-}
function occur inside it and prevent the extended environment from
mutation. However, it is still possible to mutate the extended environment
using the \c{\<\<-} function; but, that happens rarely.

@@

We look at how \emph{polymorphic} the \c{expr} argument of \eval can be, \ie
how many different types of the resolved \c{expr} there are per call sites,
in Figure~\ref{fig:polymorphism}. Most of the call sites, \ie \PercentMonomorphic,  are
\emph{consistent}. However, a few ones are \emph{polymorphic}, such as:

\begin{figure}[!h]
	\centering
    \includegraphics[width=\columnwidth]{polymorphism}
    \caption{Number of different types  of \c{expr} per call sites.} \label{fig:polymorphism}
\end{figure}
%
%\subsection{base}
%
%%There are 30 functions in \emph{base} that use \eval. Only five of them use \c{parse}, for legitimate reasons, such \c{source} and \c{sys.source} which load a file and then evaluate it, or \c{invokeRestartInteractively} which prompts the user for arguments to restart execution after an error.
%%
%%We endeavored  to write these functions without \eval. \c{match.arg} appears in \MatchArgPercent of the \eval calls. It is used to get the arguments, possibly abbreviated, of a function. With one argument, as \c{match.arg(arg)}, it uses \c{eval} to infer them from a parent function's formal arguments. We can rewrite it using its version with two arguments, \c{match.arg(arg, choices)} where \c{choices} are the possible formal arguments of the function, as shown in Listing~\ref{code:matcharg}.
%%
%%\begin{lstlisting}[caption={Rewriting \c{match.arg} without \eval}, label=code:matcharg]
%%f <- function(type=c("examples", "tests", "vignettes")) {
%%actual_type = match.arg(type)
%%...
%%}
%%
%%f <- function(type=c("examples", "tests", "vignettes")) {
%%actual_type = match.arg(type, choices=c("examples", "tests", "vignettes"))
%%...
%%}
%%\end{lstlisting}

>>>>

Richards et al.~\cite{ecoop11} provided the first large-scale study of the
runtime behavior of \eval in JavaScript. They analyzed a corpus of the
10,000 most popular websites. They used an instrumented web browser to
gather execution traces and perform a dynamic analysis.  Their results show
that \eval is pervasive with 82\% of the most popular websites using it. The
reasons for using \eval include the desire to load code on demand (to reduce
the time it takes to display a page), the deserialization of JSON data
(about 37\% of all calls) and lightweight meta-programming to customize web
pages.  While many uses \eval were legitimate, many were unnecessary and
could be replaced with equivalent and safer code.  The reported on a
categorization of inputs to \eval that covers the vast majority of input
strings.  Restricting themselves to \eval in which all named variables refer
to the global scope, many patterns can be replaced by more disciplined
code~\cite{oopsla12b}.  The limitation of that work are that there was no
measure of code coverage, so the numbers presented can be a lower bound on
the possible behaviors.  Furthermore in 2011, JavaScript usage is quite
different from today. Their work captured very few libraries and none of the
more recent Node.js ecosystem.

@@

@@

\begin{lstlisting}
> r <- 5
> env <- new.env()
> env$r <- 9
> eval(r, envir = env)
[1] 5
> evalq(r, envir = env)
[1] 9
\end{lstlisting}

@@

To roughly estimate the complexity of the code being evaluated,
Fig.~\ref{fig:ast-size} shows the distribution of the abstract syntax tree
size of the \c{expr} argument to \eval. The majority of inputs across core
and packages (\NbAstOnePercent of \eval calls and \NbAstOneCallSitePercent of call sites) have size 1. That is to say, the code being evaluated
consists of a single variable name or constant. A function call needs at
least three nodes and a list such as \c{c(1, 2, 3)} counts for 4 nodes. Inputs smaller than \AstSizeNineFive nodes account for 95\% of all
calls. Sizes in core are larger than in packages. It is due to core

@@

One of the main differences of the R \eval with the javascript one is how it
interacts with \emph{scope}, or environments in R. In Javascript, \c{eval} can
access the local scope and the global scope. In R, \c{eval} has an argument
\c{envir} to indicate in which environment its first argument must be evaluated.
By default, it is equal to \c{parent.frame()}, 
which is the parent environment of \c{eval} call. This default argument
happens in \DefaultEnvirExprPercent of the eval calls.

@@

We looked at the environments passed to all the \eval calls in our corpus.
Table~\ref{tab:environments} summarizes the results. A numeric environment class
\c{n} denotes the environment of the $n$-th call stack frame from the current
function. \c{global} denotes the top-level environment and \c{list} denotes a list
passed for evaluation of formulas. Environment classes of the form $n+$ denote
the $n$-th environment extended with a new environment. The new environment
provides a limited form of sandboxing. All assignments using the \c{\<-}
function occur inside it and prevent the extended environment from
mutation. However, it is still possible to mutate the extended environment
using the \c{\<\<-} or \c{assign} functions; but, that happens rarely.

@@

We look at how \emph{consistent} the \c{expr} argument of \eval can be, \ie
how many different types of the resolved \c{expr} there are per call sites. Most of the call sites, \ie \PercentMonomorphic,  are
\emph{consistent}, and this is similar to javascript. However, a few ones are highly \emph{polymorphic} (10 different types). They are the  pipe  operators \c{\%>\%}, \c{\%<>\%} and \c{\%$\%} in package \emph{magrittr}. It is effectively used to compose functions on their first argument, which can be of any type.

%\begin{figure}[!h]
%	\centering
%    \includegraphics[width=\columnwidth]{polymorphism}
%    \caption{Number of different types  of \c{expr} per call sites.} \label{fig:polymorphism}
%\end{figure}
%
%\subsection{base}
%
%%There are 30 functions in \emph{base} that use \eval. Only five of them use \c{parse}, for legitimate reasons, such \c{source} and \c{sys.source} which load a file and then evaluate it, or \c{invokeRestartInteractively} which prompts the user for arguments to restart execution after an error.
%%
%%We endeavored  to write these functions without \eval. \c{match.arg} appears in \MatchArgPercent of the \eval calls. It is used to get the arguments, possibly abbreviated, of a function. With one argument, as \c{match.arg(arg)}, it uses \c{eval} to infer them from a parent function's formal arguments. We can rewrite it using its version with two arguments, \c{match.arg(arg, choices)} where \c{choices} are the possible formal arguments of the function, as shown in Listing~\ref{code:matcharg}.
%%
%%\begin{lstlisting}[caption={Rewriting \c{match.arg} without \eval}, label=code:matcharg]
%%f <- function(type=c("examples", "tests", "vignettes")) {
%%actual_type = match.arg(type)
%%...
%%}
%%
%%f <- function(type=c("examples", "tests", "vignettes")) {
%%actual_type = match.arg(type, choices=c("examples", "tests", "vignettes"))
%%...
%%}
%%\end{lstlisting}
